Notes on Matrix Factorization for Hartree-Fock Stability of HEG

### 2017-04-05 doc/papers/matrixproperties/MatrixProperties.tex
Added eigenvalue histograms and updated matrixproperties.

\section{Definitions}
According to Seeger and Pople\cite{Seeger1977}, (and many other sources) the molecular orbital
Hessian has the form,
\begin{eqnarray}\label{basic}
\mathbf{H} =
  \begin{bmatrix}
    \mathbf{A}   & \mathbf{B}   \\
    \mathbf{B^*} & \mathbf{A^*} \\
  \end{bmatrix}
\end{eqnarray}
Where the matrices denoted by $\mathbf{A}$ and $\mathbf{B}$ are given by,
\begin{eqnarray}
  A_{st} &=& \e\diag +  \textcolor{myred}{\braket{aj||ib}} \\
  B_{st} &=& \textcolor{myblue}{\braket{ab||ij}}         
\end{eqnarray}
and the two electron integral used here is defined as, 
\begin{eqnarray}
  \braket{pq|rs} &=& \int\int \chi^*_p(1) \chi^*_q(2) \frac{1}{r_{12}} \chi_r(1) 
  \chi_s(2)
  d\tau d\sigma \\
  \braket{pq||rs} &=& \braket{pq|rs} - \braket{pq|sr} \\
  (pq|rs) &=& \int\int \chi^*_p(1) \chi^*_q(2) \frac{1}{r_{12}} \chi_r(1) 
  \chi_s(2)
  d\tau \\
  (pq||rs) &=& (pq|rs) - (pq|sr)
\end{eqnarray}
Where $\sigma$ is the spin coordinate and $d\tau$ is the volume element in all spatial coordinates.
Additionally, the convention of $a, b$ being virtual states, $i, j$ being occupied states and 
$p, q, r, s$ being any state is followed.

@@

\section{Forms of equations with UHF Solution}
In the case of a stationary UHF solution, the elements of matrices $\mathbf{A}$ and $\mathbf{B}$ 
have the following forms, after integrating over spin:
\begin{eqnarray}
  A_{i \rightarrow a, j \rightarrow b} &=& \e\diag \textcolor{myred}{+ 
  \delta_{\sigma_i\sigma_a}\delta_{\sigma_j\sigma_b}\Ap
                     - \delta_{\sigma_a\sigma_b}\delta_{\sigma_i\sigma_j}\App} \\
  B_{i \rightarrow a,j \rightarrow b} &=& 
  \textcolor{myblue}{\delta_{\sigma_a\sigma_i}\delta_{\sigma_b\sigma_j}\B 
             - \delta_{\sigma_a\sigma_j}\delta_{\sigma_b\sigma_i}\Br} 
\end{eqnarray}
Writing down the matrices, where the row index corresponds to $i \rightarrow a$ excitations  and 
the column index corresponds to $j \rightarrow b$ excitations:
\begin{eqnarray*}
  \mathbf{A}&=&\kbordermatrix{
        & \AtoA           & \AtoB           & \BtoA          & \BtoB          \\
  \AtoA & \e\diag + \Aa   & 0               & 0              & \Ap            \\
  \AtoB & 0               & \e\diag - \App  & 0              & 0              \\
  \BtoA & 0               & 0               & \e\diag - \App & 0              \\
  \BtoB & \Ap             & 0               & 0              & \e\diag + \Aa  
}
\end{eqnarray*}
\begin{eqnarray*}
  \mathbf{B}&=&\kbordermatrix{
        & \AtoA           & \AtoB           & \BtoA          & \BtoB          \\
  \AtoA & \Ba             & 0               & 0              & \B             \\
  \AtoB & 0               & 0               & -\Br           & 0              \\
  \BtoA & 0               & -\Br            & 0              & 0              \\
  \BtoB & \B              & 0               & 0              & \Ba            
}
\end{eqnarray*}
\\
These matrices factorize into ``spin conserved'' $(\mathbf{A', B'})$ and
``spin-unconserved'' $(\mathbf{A'', B''})$ parts, to use the language of Seeger
and Pople. Similarly, $\mathbf{H}$ factorizes, with no approximations, into
$\mathbf{H}'$ and $ \mathbf{H}''$
\begin{eqnarray}
\mathbf{H}' =
  \begin{bmatrix}
    \mathbf{A}'   & \mathbf{B}'   \\
    \mathbf{B}'^* & \mathbf{A}'^* \\
  \end{bmatrix}
\end{eqnarray}
\begin{eqnarray}
\mathbf{H}'' =
  \begin{bmatrix}
    \mathbf{A}''   & \mathbf{B}''   \\
    \mathbf{B}''^* & \mathbf{A}''^* \\
  \end{bmatrix}
\end{eqnarray}
The spin conserved matrices are given by

@@

Applying the transformation defined by the unitary matrix, 
\begin{eqnarray}
  \mathbf{U} = \frac{1}{\sqrt{2}} 
  \begin{bmatrix}
    \mathbf{1} & -\mathbf{1} & 0 &  0 \\
    \mathbf{1} &  \mathbf{1} & 0 &  0 \\
    0 &  0 & \mathbf{1} & -\mathbf{1} \\
    0 &  0 & \mathbf{1} &  \mathbf{1} \\
  \end{bmatrix}
\end{eqnarray}

With no further assumptions, this transformation doesn't do much. However, if the spatial
orbitals corresponding to $\alpha$ and $\beta$ are assumed equivalent, the 
transformed $\mathbf{H}'$ is
\begin{eqnarray*}
  \hspace{-1cm}
  \mathbf{U^{-1}H'U} =
  \begin{bmatrix}
    \e + \Aa + \Ap & 0              & \Ba + \B            & 0            \\
    0                    & \e + \Aa - \Ap    & 0                & \Ba - \B              \\
    \Ba^* + \B^*         & 0                & \e+ \Aa^* + \Ap^*   & 0                     \\
    0                    & \Ba^* - \B^*     & 0                 & \e+ \Aa^* - \Ap^*  \\
  \end{bmatrix}
\end{eqnarray*}

while the corresponding eigenvectors are 
\begin{eqnarray*}
  \mathbf{U^{-1}D'} = \frac{1}{\sqrt{2}}
  \begin{bmatrix}
    \mathbf{d_\AtoA + d_\BtoB} \\
    \mathbf{d_\AtoA - d_\BtoB} \\
    \mathbf{d^*_\AtoA + d^*_\BtoB} \\        
    \mathbf{d^*_\AtoA - d^*_\BtoB} 
  \end{bmatrix}
\end{eqnarray*}

@@

\section{Numerical result}
To verify numerically, the diagonalization of the matrices $\mathbf{H}, \mathbf{H}'$ and 
${}^3\mathbf{H}'$ were implemented independently of one another. No factorizations or 
simplifications were performed. The lowest eigenvalue of each is shown below for a constant number
of K-points and a varying density. The results are equivalent within the convergence criteria of 
Davidson's Algorithm.  
\begin{figure}[h]
  \includegraphics[width=0.75\linewidth]{../../images/Matrix_Compare.pdf}
\end{figure}

>>>>

\section{Definitions}
According to Seeger and Pople\cite{Seeger1977}, (and many other sources) the molecular orbital
Hessian has the form,
\begin{eqnarray}\label{basic}
\mathbf{H} =
  \begin{bmatrix}
    \mathbf{A}   & \mathbf{B}   \\
    \mathbf{B^*} & \mathbf{A^*} \\
  \end{bmatrix}
\end{eqnarray}
Where the matrices denoted by $\mathbf{A}$ and $\mathbf{B}$ are given by,
\begin{eqnarray}
  A_{i \rightarrow a, j \rightarrow b} &=& \e\diag +  \textcolor{myred}{\braket{aj||ib}} \\
  B_{i \rightarrow a, j \rightarrow b} &=& \textcolor{myblue}{\braket{ab||ij}}         
\end{eqnarray}
and the two electron integral used here is defined as, 
\begin{eqnarray}
  \braket{pq|rs} &=& \int\int \chi^*_p(1) \chi^*_q(2) \frac{1}{r_{12}} \chi_r(1) 
  \chi_s(2)
  d\tau d\sigma \\
  \braket{pq||rs} &=& \braket{pq|rs} - \braket{pq|sr} \\
  (pq|rs) &=& \int\int \chi^*_p(1) \chi^*_q(2) \frac{1}{r_{12}} \chi_r(1) 
  \chi_s(2)
  d\tau \\
  (pq||rs) &=& (pq|rs) - (pq|sr)
\end{eqnarray}
Where $\sigma$ is the spin coordinate and $d\tau$ is the volume element in all spatial coordinates.
Additionally, the convention of $a, b$ being virtual states, $i, j$ being occupied states and 
$p, q, r, s$ being any state is followed.

@@

\section{Forms of equations with UHF Solution}
In the case of a stationary UHF solution, the elements of matrices $\mathbf{A}$ and $\mathbf{B}$ 
have the following forms, after integrating over spin:
\begin{eqnarray}
  A_{i \rightarrow a, j \rightarrow b} &=& \e\diag \textcolor{myred}{+ 
  \delta_{\sigma_i\sigma_a}\delta_{\sigma_j\sigma_b}\Ap
                     - \delta_{\sigma_a\sigma_b}\delta_{\sigma_i\sigma_j}\App} \\
  B_{i \rightarrow a,j \rightarrow b} &=& 
  \textcolor{myblue}{\delta_{\sigma_a\sigma_i}\delta_{\sigma_b\sigma_j}\B 
             - \delta_{\sigma_a\sigma_j}\delta_{\sigma_b\sigma_i}\Br} 
\end{eqnarray}
Writing down the matrices, where the row index corresponds to $i \rightarrow a$ excitations  and 
the column index corresponds to $j \rightarrow b$ excitations:
\begin{eqnarray*}
  \mathbf{A}&=&\kbordermatrix{
        & \AtoA           & \AtoB           & \BtoA          & \BtoB          \\
  \AtoA & \e\diag + \Aa   & 0               & 0              & \Ap            \\
  \AtoB & 0               & \e\diag - \App  & 0              & 0              \\
  \BtoA & 0               & 0               & \e\diag - \App & 0              \\
  \BtoB & \Ap             & 0               & 0              & \e\diag + \Aa  
}
\end{eqnarray*}
\begin{eqnarray*}
  \mathbf{B}&=&\kbordermatrix{
        & \AtoA           & \AtoB           & \BtoA          & \BtoB          \\
  \AtoA & \Ba             & 0               & 0              & \B             \\
  \AtoB & 0               & 0               & -\Br           & 0              \\
  \BtoA & 0               & -\Br            & 0              & 0              \\
  \BtoB & \B              & 0               & 0              & \Ba            
}
\end{eqnarray*}
\\
These matrices factorize into ``spin conserved'' $(\mathbf{A', B'})$ and
``spin-unconserved'' $(\mathbf{A'', B''})$ parts, to use the language of Seeger
and Pople. Similarly, $\mathbf{H}$ factorizes, with no further approximations, into
$\mathbf{H}'$ and $ \mathbf{H}''$
\begin{eqnarray}
\mathbf{H}' =
  \begin{bmatrix}
    \mathbf{A}'   & \mathbf{B}'   \\
    \mathbf{B}'^* & \mathbf{A}'^* \\
  \end{bmatrix}
\end{eqnarray}
\begin{eqnarray}
\mathbf{H}'' =
  \begin{bmatrix}
    \mathbf{A}''   & \mathbf{B}''   \\
    \mathbf{B}''^* & \mathbf{A}''^* \\
  \end{bmatrix}
\end{eqnarray}
The spin conserved matrices are given by

@@

Applying the transformation defined by the unitary matrix, 
\begin{eqnarray}
  \mathbf{U} = \frac{1}{\sqrt{2}} 
  \begin{bmatrix}
    \mathbf{1} & -\mathbf{1} & 0 &  0 \\
    \mathbf{1} &  \mathbf{1} & 0 &  0 \\
    0 &  0 & \mathbf{1} & -\mathbf{1} \\
    0 &  0 & \mathbf{1} &  \mathbf{1} \\
  \end{bmatrix},
\end{eqnarray}
and assuming the spatial orbitals corresponding to $\alpha$ and $\beta$ are equivalent (ie RHF), 
the transformed $\mathbf{H}'$ is
\begin{eqnarray*}
  \hspace{-1cm}
  \mathbf{U^{-1}H'U} =
  \begin{bmatrix}
    \e + \Aa + \Ap & 0              & \Ba + \B            & 0            \\
    0                    & \e + \Aa - \Ap    & 0                & \Ba - \B              \\
    \Ba^* + \B^*         & 0                & \e+ \Aa^* + \Ap^*   & 0                     \\
    0                    & \Ba^* - \B^*     & 0                 & \e+ \Aa^* - \Ap^*  \\
  \end{bmatrix}
\end{eqnarray*}

while the corresponding eigenvectors are 
\begin{eqnarray*}
  \mathbf{U^{-1}D'} = \frac{1}{\sqrt{2}}
  \begin{bmatrix}
    \mathbf{d_\AtoA + d_\BtoB} \\
    \mathbf{d_\AtoA - d_\BtoB} \\
    \mathbf{d^*_\AtoA + d^*_\BtoB} \\        
    \mathbf{d^*_\AtoA - d^*_\BtoB} 
  \end{bmatrix}.
\end{eqnarray*}

@@

\section{Numerical result}
To verify numerically, the brute force diagonalization of the matrices $\mathbf{H}, \mathbf{H}'$ 
and 
${}^3\mathbf{H}'$ were implemented independently of one another. No factorizations or 
simplifications were performed. Let $\lambda_{\mathbf{M}}$ be the 
set\footnote{Technically a multiset, as a set has only unique elements.} of eigenvalues of 
square matrix $\mathbf{M}$, repeated according to their multiplicity, such that 
\begin{eqnarray*}
  \lambda_{\mathbf{H}} &=& \{ \text{All eigenvalues of } \mathbf{H}  \} \\
  \lambda_{\mathbf{H'}} &=& \{ \text{All eigenvalues of } \mathbf{H'}  \} \\
  \lambda_{\mathbf{{}^3H'}} &=& \{ \text{All eigenvalues of } \mathbf{{}^3H'}  \} \\
  \lambda_{\mathbf{{}^1H'}} &=& \{ \text{All eigenvalues of } \mathbf{{}^1H'}  \} 
\end{eqnarray*}
To machine precision, the calculated eigenspectrum of  $\mathbf{H}'$ is
equivalent to the union of the eigenvalues of ${}^3\mathbf{H}'$ and the eigenvalues of
${}^1\mathbf{H}'$.
\begin{equation}
  \lambda_{\mathbf{H}'} = \lambda_{\mathbf{{}^3H}'} \cup \lambda_{\mathbf{{}^1H}'}  
\end{equation}
Similarly, the calculated eigenspectrum of $\mathbf{H}$ is equivalent to
the union of the eigenvalues of ${}^1\mathbf{H}'$ and thrice-repeated eigenvalues
of ${}^3\mathbf{H}'$ (two coming from the RHF limit of $\mathbf{H}''$ and one coming from
$\mathbf{H}'$)
\begin{equation}
  \lambda_{\mathbf{H}} = \lambda_{\mathbf{{}^3H}'} \cup
                              \lambda_{\mathbf{{}^3H}'} \cup
                              \lambda_{\mathbf{{}^3H}'} \cup
                               \lambda_{\mathbf{{}^1H}'}
\end{equation}
Visually, this information is presented in figure 1. This shows numerically that the RHF-GHF 
instability is indistinguishable from the RHF-UHF instability, given an RHF wavefunction. 
Lifting the symmetry of spatial orbitals between $\alpha$ and $\beta$ spin states will remove the
degeneracies seen here, and the resulting UHF-GHF instability will become distinct from the RHF-UHF
instability.  
\begin{figure}[h]\label{histograms}
  \includegraphics[width=\linewidth]{../../images/matrix_factor_histograms.pdf}
  \caption{ Histogram of the calculated eigenspectra of stability matrices and factorizations. 
  \newline \textbf{[Left]} The eigenspectrum of $\mathbf{H}$ (black outline, no fill) is equivalent 
  to the 
  set formed by combining the eigenspectra of $\mathbf{{}^3H'}$ (green), repeated thrice, and 
  $\mathbf{{}^1H'}$ 
  (blue).
  \newline \textbf{[Right]} The eigenspectrum of $\mathbf{H'}$ (black outline, no fill) is 
  equivalent to the 
  set formed by combining the eigenspectra of $\mathbf{{}^3H'}$ (green) and $\mathbf{{}^1H'}$ 
  (blue).  
  \newline Although not apparent due to binning, all eigenvalues of factorized matrices agree
  with eigenvalues of the unfactored matrix to machine precision. 
  }
\end{figure}
